---
title: "Guide_Info_formatting"
output: html_document
date: "2023-11-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# About

Given the genomic coordinates of the guides from the Klann et al. paper, we align these coordinates up with what we've determined as CREs and format a final table for the CRISPRi pipeline

```{r, message=FALSE}
library(readxl)
library(Seurat)
library(GenomicRanges)
library(dplyr)
library(readr)
library(rtracklayer)
library(tidyverse)
```

# Import Files

Import the candidate cres file

```{r}
# Import the bed file of DHS sites
candidate_cres <- read.table("../candidate_cre_data/sample1_candidate_cres.bed", header=FALSE, sep="\t", stringsAsFactors=FALSE, quote="")
# Keep the first four columns
candidate_cres <- candidate_cres[,c(1:4)]
# Rename the columns
colnames(candidate_cres) <- c("target_chr", "target_start", "target_end", "target_name")
```

Import the excel file with all of the guide information

```{r}
klann_guide_info <- read_xlsx("supplementary_table_16_scCERES_library_gRNAs.xlsx")
```

# Liftover Klann Coordinates

The following is optional as the output is provided as `gr_object_final_df.csv` which is loaded below. The Klann guide coordinates are in hg19, so we lift them over to hg38 using `easylift` (v1.0.0). In line with the other R Markdown files, the analysis of Klann et al. 2021 uses R version 4.1.2, but `easylift` is only compatible with R versions 4.3+, so the following code block was uesd to perform `easylift` in an environment using R version 4.3. The chain file used in this code block can be downloaded from : `https://hgdownload.cse.ucsc.edu/goldenPath/hg19/liftOver/hg19ToHg38.over.chain.gz`

```{r}
# library(GenomicRanges)
# library(easylift)
# library(rtracklayer)
# library(readxl)
# library(dplyr)
# library(stringr)
# 
# 
# # Load the data
# klann_guide_info <- read_xlsx("../Paper/GSE167890_supplementary_tables/supplementary_table_16_scCERES_library_gRNAs.xlsx")
# 
# # Keep the necessary columns
# klann_guide_info <- klann_guide_info %>%
#   filter(!is.na(gRNA_start))
# 
# # Convert to GRanges
# # Function to load a BED file into a GRanges object
# gr_object <- GRanges(
#   seqnames = klann_guide_info$chrom,
#   ranges = IRanges(start = klann_guide_info$gRNA_start, end = klann_guide_info$gRNA_end),
#   mcols = data.frame(chrom = klann_guide_info$chrom,
#                      DHS_annotation = klann_guide_info$DHS_annotation,
#                      gRNA_start = klann_guide_info$gRNA_start,
#                      gRNA_end = klann_guide_info$gRNA_end,
#                      gRNA_ID = klann_guide_info$gRNA_ID,
#                      protospacer = klann_guide_info$protospacer,
#                      gRNA_strand = klann_guide_info$gRNA_strand
#                      )
# )
# 
# genome(gr_object) <- "hg19"
# 
# # LiftOver
# # Assuming the chain file for hg19 to hg38 is in the working directory
# chain <- system.file("extdata", "hg19ToHg38.over.chain.gz", package = "easylift")
# 
# # LiftOver the GRanges objects from hg19 to hg38 using easylift
# gr_object_lifted <- easylift(gr_object, to = "hg38", chain)
# 
# # Output the lifted GRanges objects
# gr_object_lifted
# 
# # Export the files
# # First lets remove any guides which have mapped to a shortened sequence
# gr_object_final_df <- as.data.frame(gr_object_lifted)[as.data.frame(gr_object_lifted)$width == 20,]
# 
# # Let's update the column names 
# new_cols <- c("seqnames", "hg38_start", "hg38_end", "width", "strand", 
#               "chrom", "DHS_annotation", "gRNA_start", "gRNA_end", 
#               "gRNA_ID", "protospacer", "gRNA_strand")
# colnames(gr_object_final_df) <- new_cols
# 
# # These are the gRNA IDs that didn't lift over
# print("These are the gRNA IDs that didn't lift over")
# setdiff(klann_guide_info$gRNA_ID, as.data.frame(gr_object_lifted)$mcols.gRNA_ID)
# 
# print("These are the gRNA IDs that don't have a protospacer width of 20 after liftover")
# setdiff(as.data.frame(gr_object_lifted)$mcols.gRNA_ID, gr_object_final_df$gRNA_ID)
# 
# print("These are the gRNA IDs that don't have a protospacer width of 20 after liftover")
# as.data.frame(gr_object_lifted)[as.data.frame(gr_object_lifted)$width != 20, "mcols.gRNA_ID"]
```

```{r}
# Load in the object
gr_object_final_df <- read.csv("gr_object_final_df.csv", header = TRUE)
```

# Reformat DF

## Rename Columns

```{r}
columns_to_keep <- c("seqnames", "hg38_start", "hg38_end", "DHS_annotation", 
                     "gRNA_ID", "protospacer", "gRNA_strand")
guide_info <- gr_object_final_df[,columns_to_keep]
colnames(guide_info) <- c("chr", "start", "end", "target_type", "name", "spacer", "strand")

# If the target_type begins with "Promoter", label it as TSSCtrl, if not, label it as enh
guide_info <- guide_info %>%
  mutate(target_type = ifelse(grepl("^Promoter", target_type, ignore.case = TRUE), "TSSCtrl", "enh"))
```

## Sanity Checks

```{r}
# Print the number of promoters in the original and filtered datasets
print("The number of annotations that begin with 'Promoter' in the original dataset")
length(klann_guide_info$DHS_annotation[grepl("^Promoter", klann_guide_info$DHS_annotation)])
print("The number of annotations that begin with 'Promoter' in the filtered dataset")
length(guide_info$target_type[grepl("^TSSCtrl", guide_info$target_type)])

# Print the number of non-targeting guides
print("The number of non-targeting guides")
length(klann_guide_info$DHS_annotation[is.na(klann_guide_info$DHS_annotation)])

# Print the number of guides that don't target Promoters
print("The number of guides targeting enhancer regions in the original dataset")
length(klann_guide_info$DHS_annotation[!grepl("^Promoter", klann_guide_info$DHS_annotation, ignore.case = TRUE) & !is.na(klann_guide_info$DHS_annotation)])
print("The number of guides targeting enhancer regions in the filtered dataset")
length(guide_info$target_type[!grepl("^TSSCtrl", guide_info$target_type, ignore.case = TRUE)])

# Print the total guides
print("The total number of guides we now have")
length(guide_info$chr)
```

This all makes sense because 7 guides were removed.

# Mapping Guides to CREs

```{r}
# Create the GRanges objects
guides_gr <- makeGRangesFromDataFrame(
  guide_info[,c("chr", "start", "end")], 
  starts.in.df.are.0based = TRUE)

targets_gr <- makeGRangesFromDataFrame(
  candidate_cres[,c("target_chr", "target_start", "target_end")], 
  starts.in.df.are.0based = TRUE)


# Overlap the GRanges objects and create the `guide_targets` table
overlaps <- findOverlaps(query = guides_gr, subject = targets_gr, ignore.strand = TRUE)
guide_targets <- bind_cols(
  guide_info[queryHits(overlaps), ], 
  candidate_cres[subjectHits(overlaps), ]
  )

# Format the guide_targets file
guide_targets$target_strand <- "."
guide_targets <- guide_targets[,c("chr", "start", "end", "name", "strand", "spacer", "target_chr", "target_start", "target_end", "target_name", "target_strand", "target_type")]

print(paste0("The proportion of mapped guides are 1833/3044 or ", formatC(1833/3044 * 100, format = "f", digits = 2), "%"))
```

## Mapped Guides Stats

```{r}
# Count the number of occurrences
before_TSSCtrl <- sum(guide_info["target_type"] == "TSSCtrl")
before_enh <- sum(guide_info["target_type"] == "enh")
after_TSSCtrl <- sum(guide_targets["target_type"] == "TSSCtrl")
after_enh <- sum(guide_targets["target_type"] == "enh")

# Create a data frame for plotting
plot_data <- data.frame(
  Category = c("TSSCtrl", "TSSCtrl", "Enh", "Enh"),
  Count = c(before_TSSCtrl, after_TSSCtrl, before_enh, after_enh),
  Type = c("Before", "After", "Before", "After")
)

# Load ggplot2 for plotting
library(ggplot2)

# Create the bar plot with values displayed
ggplot(plot_data, aes(x = Category, y = Count, fill = Type, label = Count)) +
  geom_bar(stat = "identity", position = "identity") +
  geom_text(position = position_identity(), vjust = -0.5) +
  scale_fill_manual(values = c("Before" = "gray", "After" = "blue")) +
  theme_minimal() +
  labs(title = "Guide Counts Before and After Mapping", x = "Category", y = "Count")
```

# Exporting Guides

```{r}
# Save the tsv file
write_tsv(guide_targets, "results/guide_targets.tsv")
```

